Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-21 paper/paper.Rmd
Adding references and section about CRAN/GitHub

One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total. 
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

@@

@@

<!-- MH: Add references for all packages -->

<!-- JN: maybe we should also add an introduction to R packages here-->
<!-- for example, CRAN vs GitHub vs other -->
<!-- who creates R packages? -->
<!-- who maintains them and what is an archived package -->
<!-- resources on how to create your own packages - https://r-pkgs.org/ -->
<!-- comment on never-ending changes in r world -->
<!-- MH: I like this idea. I will add a short paragraph -->

## Spatial data representations {#spatrep}

@@

<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
<!-- rgee -->

@@

All of the packages listed in Section \@ref(spatrep) have build-in methods for plotting spatial data using the `plot()` function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library.<!--this sentence can be easily removed-->
R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package. 
Of course, also packages specifically designed for SDM exists. 
This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
<!-- JN: I think all of the above packages should be cited -->
<!-- MH: Yeah, will do. I was just to lazy  -->

>>>>

One example of a successful open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total. 
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

@@

There are two major platforms to host *R* packages and make them easily accessible to potential users. Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
Additionally, hosting the package on *GitHub* provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
However, at the same time, releasing a *R* package on *CRAN* requires more initial work for the developers compared to *GitHub*.
Since most *R* packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many *R* packages depend on each other (i.e. package *x* requires and uses code from package *y*) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
For more information about *R* package development, see \cite{Wickham2015}.

@@

## Spatial data representations {#sec:spatrep}

@@

<!-- JN: I am also thinking (unsure yet) about mentioning R as a bridge between other software (e.g. GRASSGIS) -->
<!-- rgee -->
<!-- MH: Maybe one short sentence that R allows to connect to other open-source software. We already introduce the topic previously anyhow -->

@@

All of the packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the `plot()` function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library.<!--this sentence can be easily removed-->
R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
Popular approaches and packages include generalized linear models using, e.g., the *stats* \cite{RCoreTeam2019} package; generalized additive models using, e.g., the *mgcv* \cite{Wood2017} or *lme4* \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the *rpart* \cite{Therneau2019}, *randomForest* \cite{Liaw2002} or *ranger* \cite{Wright2017} package or multivariate data analysis using , e.g., the *ade4* \cite{Dray2007} or *vegan* \cite{Oksanen2019} package. 
Of course, also packages specifically designed for SDM exists. 
This includes the *dismo* \cite{Hijmans2017}, *sdm* \cite{Naimi2016}, *ecospat* \cite{Broennimann2020}, *biomod2* \cite{Thuiller2020} and *PresenceAbsence* \cite{Freeman2008} packages.

### 2020-07-21 paper/paper.tex
Adding references and section about CRAN/GitHub

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
     \email{\href{mailto:mhk.hesselbarthh@gmail.com}{\nolinkurl{mhk.hesselbarthh@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Author 3 \at
     Adress author 3 \\
     \email{E-Mail author 3}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        \ldots{} \at
     \ldots{} \\
     \email{\ldots{}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

One example of a successful open-source project is the \emph{R} programming language and its \emph{Comprehensive R Archive Network} (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even much higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

@@

\hypertarget{spatrep}{%
\subsection{Spatial data representations}\label{spatrep}}

@@

All of the packages listed in Section \ref{spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} package; generalized additive models using, e.g., the \emph{mgcv} or \emph{lme4} package; classification and regression trees (CART) using, e.g., the \emph{rpart}, \emph{randomForest} or \emph{ranger} package or multivariate data analysis using , e.g., the \emph{ade4} or \emph{vegan} package.
Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and \emph{PresenceAbsence} packages.

>>>>

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
     \email{\href{mailto:mhk.hesselbarth@gmail.com}{\nolinkurl{mhk.hesselbarth@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Author 3 \at
     Adress author 3 \\
     \email{E-Mail author 3}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        \ldots{} \at
     \ldots{} \\
     \email{\ldots{}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

One example of a successful open-source project is the \emph{R} programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even much higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

@@

There are two major platforms to host \emph{R} packages and make them easily accessible to potential users. Firstly, hosting the package on \emph{GitHub} allows users to install it with only one line of code.
Additionally, hosting the package on \emph{GitHub} provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
However, at the same time, releasing a \emph{R} package on \emph{CRAN} requires more initial work for the developers compared to \emph{GitHub}.
Since most \emph{R} packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many \emph{R} packages depend on each other (i.e.~package \emph{x} requires and uses code from package \emph{y}) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
For more information about \emph{R} package development, see \cite{Wickham2015}.

\hypertarget{sec:spatrep}{%
\subsection{Spatial data representations}\label{sec:spatrep}}

@@

All of the packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the \texttt{plot()} function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library.
R's general-purpose plotting package \emph{ggplot2} together with its extension \emph{ggspatial} also offers some capabilities for plotting spatial data.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020} and \emph{PresenceAbsence} \cite{Freeman2008} packages.
